---
title: "MultiAncestryPGxGWAS: Simulation and Application"
author: "Wujuan Zhong, Judong Shen" 
date: "`r Sys.Date()`"
output: 
  html_document: default
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{MultiAncestryPGxGWAS: Simulation and Application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

MultiAncestryPGxGWAS is an R package to conduct genome-wide genotype-by-environment (GxE) or genotype-by-treatment interaction (GxT) analysis powered by the following methods:
Linear regression with robust (sandwich) variance, 

This vignette demonstrates how to conduct genome-wide GxT analysis using an example genotype data in PLINK v1 binary format (bed/bim/fam).

- Generate or load example genotype data in PLINK v1 binary format.
- Then separately for continuous and binary outcomes, 
  - Simulate phenotypes and covariates.
  - Run genome-wide GxT association.
- Summarize the results.

The preparation and simulation examples use a simulated genotype data; association analyses require the external tools listed below.

- PLINK1.9 (https://www.cog-genomics.org/plink/1.9/): v1.90b5.1
- PLINK2 (https://www.cog-genomics.org/plink/2.0/): v2.00a2LM
- GCTA (https://yanglab.westlake.edu.cn/software/gcta/#Download): v1.95.3
- LEMMA (https://github.com/mkerin/LEMMA.git): 1.0.2+
- GEM (https://github.com/large-scale-gxe-methods/GEM.git): 1.4.5+
- bgenix (https://enkre.net/cgi-bin/code/bgen/dir?ci=trunk): 1.1.7
- MR-MEGA (https://genomics.ut.ee/en/tools): 0.2
- (optional) mpirun (https://docs.open-mpi.org/en/main/index.html): for MPI-based LEMMA runs: 4.1.7

# Apply MultiAncestryPGxGWAS to a Continuous Outcome and a Binary Outcome

- Continuous outcome methods:
  `fastGWA_GE`, `LEMMA`, `GENESIS`, `LR-Sandwich`, fixed-effects meta-analysis
  of `LR-Sandwich`, and `MR-MEGA` of `LR-Sandwich`.
- Binary outcome methods:
  `fastGWA_GE`, `LEMMA`, `GENESIS`, `GEM`, and `MR-MEGA` of `GEM`.
  Note: `fastGWA_GE` and `LEMMA` naively treat the binary outcome as `0/1`. 

These examples require external tools (`plink`, `plink2`, `flashpca`, `gcta`, `lemma_1_0_4`,
  `GEM`, `bgenix`, `MR-MEGA`) plus `GENESIS` R package.

## We first set up paths to external tools and working directories.

```{r setup-workflow, eval = TRUE}
library(MultiAncestryPGxGWAS)

# If your site does not expose the command names below,
# define `configs` with explicit executable paths.
configs <- list(
  plink1.9.path = "plink", # path to plink
  plink2.path = "plink2", # path to plink2
  flashpca.path = "flashpca_x86-64", # path to flashpca
  gcta.path = "gcta", # path to gcta
  lemma.path = "lemma_1_0_4", # path to lemma
  gem.path = "GEM_1.4.5_static", # path to GEM
  bgenix.path = "bgenix", # path to bgenix
  mrmega.path = "MR-MEGA", # path to MR-MEGA
  mpirun.path = "mpirun" # path to mpirun
)

mpgx_validate_tool_configs(configs)

work_dir <- file.path(tempdir(), "MultiAncestryPGxGWAS_vignette")

methods_output_dir <- file.path(work_dir, "model_results")
dir.create(methods_output_dir, recursive = TRUE, showWarnings = FALSE)

method_out_dirs <- list(
  genotype_artifacts = file.path(work_dir, "genotype_artifacts"),
  continuous = list(
    root = file.path(methods_output_dir, "continuous"),
    lr_sandwich = file.path(methods_output_dir, "continuous", "lr_sandwich"),
    fe_meta_lr = file.path(methods_output_dir, "continuous", "fe_meta_lr"),
    mrmega_lr = file.path(methods_output_dir, "continuous", "mrmega_lr"),
    fastgwa_ge = file.path(methods_output_dir, "continuous", "fastgwa_ge"),
    genesis = file.path(methods_output_dir, "continuous", "genesis"),
    lemma = file.path(methods_output_dir, "continuous", "lemma")
  ),
  binary = list(
    root = file.path(methods_output_dir, "binary"),
    fastgwa_ge_naive = file.path(methods_output_dir, "binary", "fastgwa_ge_naive"),
    lemma_naive = file.path(methods_output_dir, "binary", "lemma_naive"),
    genesis = file.path(methods_output_dir, "binary", "genesis"),
    gem = file.path(methods_output_dir, "binary", "gem"),
    mrmega_gem = file.path(methods_output_dir, "binary", "mrmega_gem")
  )
)

method_leaf_dirs <- unique(c(
  method_out_dirs$genotype_artifacts,
  unlist(method_out_dirs$continuous, use.names = FALSE),
  unlist(method_out_dirs$binary, use.names = FALSE)
))
invisible(lapply(method_leaf_dirs, dir.create, recursive = TRUE, showWarnings = FALSE))

```

## Prepare a small example genotype dataset used by all methods below.

We use package helper mpgx_prepare_ancestry_keep_files() for ancestry-specific keep files.

```{r prepare-genotype-data, eval = TRUE, results = 'hide'}
prepare_geno <- mpgx_prepare_example_genotype(
  output_dir = file.path(work_dir, "genotype_data"),
  prefix = "toyData",
  source = "toy_data", 
  configs = configs,
  overwrite = TRUE,
  pcs_output_dir = method_out_dirs$genotype_artifacts
)

genotype_artifacts <- mpgx_prepare_genotype_artifacts(
  genotype_prefix_or_path = prepare_geno$prefix,
  out_dir = method_out_dirs$genotype_artifacts,
  configs = configs
)

simulation_seed <- 100001
```

## Simulate one continuous-outcome run and prepare run-specific input files, reusing genotype artifacts.

```{r simulate-continuous-data, eval = TRUE, results = 'hide'}
cont <- mpgx_simulate_phenotypes(
  bfile_prefix = prepare_geno$prefix,
  ancestry_info_file = prepare_geno$ancestry_info_file,
  output_dir = file.path(work_dir, "simulated_data"),
  result_label = "continuous",
  seed = simulation_seed,
  num_pcs = 10,
  pcs_file = prepare_geno$pcs_file,
  variance_proportion = "0.25_0.25_0.05_0.05_0.00_0.40",
  n_snps_maineffects_only = 2,
  n_snps_interaction_only = 2,
  n_snps_both = 6,
  effect_size_mean = 6,
  type_of_phenotype = "normal",
  configs = configs
)

cont_keep <- mpgx_prepare_ancestry_keep_files(
  ancestry_info_file = prepare_geno$ancestry_info_file,
  out_dir = method_out_dirs$continuous$root
)

cont_inputs <- list(
  bfile_prefix = prepare_geno$prefix,
  seed = cont$seed,
  result_label = cont$result_label,
  result_dir = cont$result_dir,
  phenotype_file = cont$phenotype_file,
  qcovar_file = cont$qcovar_file,
  covar_file = cont$covar_file,
  envir_file = cont$envir_file,
  sparse_grm_prefix = genotype_artifacts$sparse_grm_prefix,
  bgen_file = genotype_artifacts$bgen_file,
  gds_file = genotype_artifacts$gds_file,
  keep_files = cont_keep$keep_files,
  lemma_pheno = file.path(cont$result_dir, 
    sprintf("sim_phenotype_forLEMMA_seed_%d.txt", cont$seed)),
  lemma_envir = file.path(cont$result_dir, "sim_envir_forLEMMA.txt"),
  lemma_covar = file.path(cont$result_dir, "sim_covar_qcovar_forLEMMA.txt"),
  gem_pheno = file.path(cont$result_dir, 
    sprintf("sim_phenotype_forGEM_seed_%d.txt", cont$seed))
)
```

## Run continuous-outcome methods. 
Each method is separated into its own chunk and stores a parsed GWAS data frame in each requested output object.

### We first run LR-Sandwich on all samples and then per ancestry to create inputs for meta-analysis methods.

```{r run-continuous-lr-sandwich, eval = TRUE, results = 'hide'}
# LR-Sandwich on all samples.
lr_sandwich_all_prefix <- file.path(
  method_out_dirs$continuous$lr_sandwich, "continuous.LRallSandwich")
lr_sandwich_all <- mpgx_run_lr_sandwich(
  bfile_prefix = cont_inputs$bfile_prefix,
  phenotype_file = cont_inputs$phenotype_file,
  qcovar_file = cont_inputs$qcovar_file,
  covar_file = cont_inputs$covar_file,
  envir_file = cont_inputs$envir_file,
  output_prefix = lr_sandwich_all_prefix,
  sandwich = TRUE,
  configs = configs
)

# LR-Sandwich per ancestry (inputs for fixed-effects meta and MR-MEGA).
ancestry_levels <- names(cont_inputs$keep_files)
lr_sandwich_by_eth <- lapply(ancestry_levels, function(ances) {
  out_prefix <- file.path(
    method_out_dirs$continuous$lr_sandwich, paste0("continuous.LRallSandwich.", ances))
  mpgx_run_lr_sandwich(
    bfile_prefix = cont_inputs$bfile_prefix,
    phenotype_file = cont_inputs$phenotype_file,
    qcovar_file = cont_inputs$qcovar_file,
    covar_file = cont_inputs$covar_file,
    envir_file = cont_inputs$envir_file,
    keep_file = cont_inputs$keep_files[[ances]],
    output_prefix = out_prefix,
    sandwich = TRUE,
    configs = configs
  )
})
names(lr_sandwich_by_eth) <- ancestry_levels
lr_sandwich_by_eth_files <- extract_gwas_files(lr_sandwich_by_eth)
```

### We then combine per-ancestry LR-Sandwich outputs using fixed-effect meta-analysis and MR-MEGA.

```{r run-continuous-meta-and-mrmega, eval = TRUE, results = 'hide'}
# Fixed-effects meta-analysis of LR-Sandwich.
meta_lr_cont_file <- file.path(
  method_out_dirs$continuous$fe_meta_lr, "continuous.meta_LRallSandwich.txt")
meta_lr_cont <- mpgx_run_fe_meta_for_lr(
  ancestry_result_files = lr_sandwich_by_eth_files,
  output_file = meta_lr_cont_file,
  ancestry_labels = ancestry_levels
)

# MR-MEGA of LR-Sandwich.
mrmega_lr_cont <- mpgx_run_mrmega_for_lr(
  ancestry_result_files = lr_sandwich_by_eth_files,
  output_prefix_base = file.path(
    method_out_dirs$continuous$mrmega_lr, "continuous.MRMEGA_LRallSandwich"),
  configs = configs
)
```

### Next, run fastGWA_GE for the continuous phenotype using the precomputed sparse GRM.

```{r run-continuous-fastgwa-ge, eval = TRUE, results = 'hide'}
# fastGWA_GE (continuous).
fastgwa_ge_cont_prefix <- file.path(
  method_out_dirs$continuous$fastgwa_ge, "continuous.fastGWA_GE")
fastgwa_ge_cont <- mpgx_run_fastgwa_ge(
  bfile_prefix = cont_inputs$bfile_prefix,
  grm_sparse_prefix = cont_inputs$sparse_grm_prefix,
  phenotype_file = cont_inputs$phenotype_file,
  qcovar_file = cont_inputs$qcovar_file,
  covar_file = cont_inputs$covar_file,
  envir_file = cont_inputs$envir_file,
  output_prefix = fastgwa_ge_cont_prefix,
  sandwich = TRUE,
  configs = configs
)
```

### Then run GENESIS on the same continuous-outcome inputs for comparison.

```{r run-continuous-genesis, eval = TRUE, results = 'hide'}
# GENESIS (continuous).
genesis_cont_file <- file.path(
  method_out_dirs$continuous$genesis, "continuous.GENESIS.txt")
genesis_cont <- mpgx_run_genesis(
  gds_file = cont_inputs$gds_file,
  sparse_grm_prefix = cont_inputs$sparse_grm_prefix,
  phenotype_file = cont_inputs$phenotype_file,
  qcovar_file = cont_inputs$qcovar_file,
  covar_file = cont_inputs$covar_file,
  envir_file = cont_inputs$envir_file,
  output_file = genesis_cont_file
)
```

### Finally, run LEMMA for continuous outcome analysis with the LEMMA-formatted inputs.

```{r run-continuous-lemma, eval = TRUE, results = 'hide'}
# LEMMA (continuous).
lemma_cont_prefix <- file.path(method_out_dirs$continuous$lemma, "continuous.LEMMA")
lemma_cont <- mpgx_run_lemma(
  bgen_file = cont_inputs$bgen_file,
  phenotype_file = cont_inputs$lemma_pheno,
  envir_file = cont_inputs$lemma_envir,
  covar_file = cont_inputs$lemma_covar,
  output_prefix = lemma_cont_prefix,
  n_tasks = 1,
  configs = configs
)
```


## Simulate one binary-outcome run and prepare run-specific local inputs, reusing genotype artifacts.

```{r simulate-binary-data, eval = TRUE, results = 'hide'}
binary <- mpgx_simulate_phenotypes(
  bfile_prefix = prepare_geno$prefix,
  ancestry_info_file = prepare_geno$ancestry_info_file,
  output_dir = file.path(work_dir, "simulated_data"),
  result_label = "binary",
  seed = simulation_seed,
  num_pcs = 10,
  pcs_file = prepare_geno$pcs_file,
  variance_proportion = "0.25_0.25_0.05_0.05_0.00_0.40",
  n_snps_maineffects_only = 2,
  n_snps_interaction_only = 2,
  n_snps_both = 6,
  effect_size_mean = 6,
  type_of_phenotype = "binary",
  configs = configs
)

bin_keep <- mpgx_prepare_ancestry_keep_files(
  ancestry_info_file = prepare_geno$ancestry_info_file,
  out_dir = method_out_dirs$binary$root
)

bin_inputs <- list(
  bfile_prefix = prepare_geno$prefix,
  seed = binary$seed,
  result_label = binary$result_label,
  result_dir = binary$result_dir,
  phenotype_file = binary$phenotype_file,
  qcovar_file = binary$qcovar_file,
  covar_file = binary$covar_file,
  envir_file = binary$envir_file,
  sparse_grm_prefix = genotype_artifacts$sparse_grm_prefix,
  bgen_file = genotype_artifacts$bgen_file,
  gds_file = genotype_artifacts$gds_file,
  keep_files = bin_keep$keep_files,
  lemma_pheno = file.path(
    binary$result_dir, sprintf("sim_phenotype_forLEMMA_seed_%d.txt", binary$seed)),
  lemma_envir = file.path(binary$result_dir, "sim_envir_forLEMMA.txt"),
  lemma_covar = file.path(binary$result_dir, "sim_covar_qcovar_forLEMMA.txt"),
  gem_pheno = file.path(
    binary$result_dir, sprintf("sim_phenotype_forGEM_seed_%d.txt", binary$seed))
)
```

## Run binary-outcome methods in separate chunks, including per-ancestry GEM preparation for MR-MEGA.

### We first run fastGWA_GE while naively treating the binary phenotype as 0/1.

```{r run-binary-fastgwa-ge-naive, eval = TRUE, results = 'hide'}
# Naively treat binary phenotype as 0/1 for fastGWA_GE.
fastgwa_bin_naive_prefix <- file.path(
  method_out_dirs$binary$fastgwa_ge_naive, "binary.fastGWA_GE_naive")
fastgwa_bin_naive <- mpgx_run_fastgwa_ge(
  bfile_prefix = bin_inputs$bfile_prefix,
  grm_sparse_prefix = bin_inputs$sparse_grm_prefix,
  phenotype_file = bin_inputs$phenotype_file,
  qcovar_file = bin_inputs$qcovar_file,
  covar_file = bin_inputs$covar_file,
  envir_file = bin_inputs$envir_file,
  output_prefix = fastgwa_bin_naive_prefix,
  sandwich = TRUE,
  configs = configs
)
```

### Next, run LEMMA with the same naive binary encoding and LEMMA-specific inputs.

```{r run-binary-lemma-naive, eval = TRUE, results = 'hide'}
# Naively treat binary phenotype as 0/1 for LEMMA.
lemma_bin_naive_prefix <- file.path(
  method_out_dirs$binary$lemma_naive, "binary.LEMMA_naive")
lemma_bin_naive <- mpgx_run_lemma(
  bgen_file = bin_inputs$bgen_file,
  phenotype_file = bin_inputs$lemma_pheno,
  envir_file = bin_inputs$lemma_envir,
  covar_file = bin_inputs$lemma_covar,
  output_prefix = lemma_bin_naive_prefix,
  configs = configs
)
```

### Then run GENESIS for binary-outcome association testing using the GDS input.

```{r run-binary-genesis, eval = TRUE, results = 'hide'}
genesis_bin_file <- file.path(method_out_dirs$binary$genesis, "binary.GENESIS.txt")
genesis_bin <- mpgx_run_genesis(
  gds_file = bin_inputs$gds_file,
  sparse_grm_prefix = bin_inputs$sparse_grm_prefix,
  phenotype_file = bin_inputs$phenotype_file,
  qcovar_file = bin_inputs$qcovar_file,
  covar_file = bin_inputs$covar_file,
  envir_file = bin_inputs$envir_file,
  output_file = genesis_bin_file
)
```

### After that, run GEM on the full sample to produce a binary-outcome GWAS result.

```{r run-binary-gem, eval = TRUE, results = 'hide'}
gem_bin_prefix <- file.path(method_out_dirs$binary$gem, "binary.GEM")
gem_bin <- mpgx_run_gem(
  bfile_prefix = bin_inputs$bfile_prefix,
  phenotype_file = bin_inputs$gem_pheno,
  output_prefix = gem_bin_prefix,
  configs = configs
)
```

### Finally, prepare per-ancestry GEM outputs and run MR-MEGA on those ancestry-specific results.

```{r run-binary-mrmega-from-gem, eval = TRUE, results = 'hide'}
# Per-ancestry GEM inputs and runs for MR-MEGA.
gem_by_eth_files <- mpgx_prepare_mrmega_gem_inputs(
  bfile_prefix = bin_inputs$bfile_prefix,
  gem_phenotype_file = bin_inputs$gem_pheno,
  keep_files = bin_inputs$keep_files,
  output_dir = method_out_dirs$binary$gem,
  configs = configs
)

# MR-MEGA of GEM (binary, OR-scale input).
pc_num_bin <- max(length(gem_by_eth_files) - 3, 0)
mrmega_gem <- mpgx_run_mrmega_for_gem(
  ancestry_result_files = gem_by_eth_files,
  output_prefix_base = file.path(method_out_dirs$binary$mrmega_gem, "binary.MRMEGA_GEM"),
  configs = configs
)
```

# Summarize Parsed GWAS Results

This chunk applies `mpgx_summarize_association()` to each method's parsed
GWAS data frame (`$gwas_df`) for both continuous and binary outcomes.

```{r summarize-parsed-results, eval = TRUE}
summarize_method_list <- function(
  method_results,
  causal_snps,
  scenario,
  seed,
  mrmega_methods = character(),
  mrmega_pc_num = NULL
) {
  summaries <- lapply(names(method_results), function(method_name) {
    method_df <- mpgx_get_method_gwas_df(method_results[[method_name]], label = method_name)
    mrmega_pc <- if (method_name %in% mrmega_methods) mrmega_pc_num else NULL
    # Keep rows when a method has no 2-df test (for example MR-MEGA or LEMMA).
    has_non_missing_2df <- "P_2df" %in% colnames(method_df) && any(!is.na(method_df$P_2df))
    summary_df <- mpgx_summarize_association(
      result = method_df,
      causal_snps = causal_snps,
      rm_na_2df = has_non_missing_2df,
      mrmega_pc_num = mrmega_pc
    )

    cbind(
      data.frame(
        method = method_name,
        scenario = scenario,
        seed = seed,
        stringsAsFactors = FALSE
      ),
      summary_df
    )
  })

  out <- do.call(rbind, summaries)
  rownames(out) <- NULL
  out
}

continuous_methods <- list(
  LRallSandwich = lr_sandwich_all,
  meta_LRallSandwich = meta_lr_cont,
  MRMEGA_LRallSandwich = mrmega_lr_cont,
  fastGWA_GE = fastgwa_ge_cont,
  GENESIS = genesis_cont,
  LEMMA = lemma_cont
)

continuous_summary <- summarize_method_list(
  method_results = continuous_methods,
  causal_snps = union(cont$causal_main_snps, cont$causal_interaction_snps),
  scenario = cont$result_label,
  seed = cont_inputs$seed,
  mrmega_methods = "MRMEGA_LRallSandwich",
  mrmega_pc_num = max(length(lr_sandwich_by_eth_files) - 3, 0)
)

continuous_summary

binary_methods <- list(
  fastGWA_GE_naive = fastgwa_bin_naive,
  LEMMA_naive = lemma_bin_naive,
  GENESIS = genesis_bin,
  GEM = gem_bin,
  MRMEGA_GEM = mrmega_gem
)

binary_summary <- summarize_method_list(
  method_results = binary_methods,
  causal_snps = union(binary$causal_main_snps, binary$causal_interaction_snps),
  scenario = binary$result_label,
  seed = bin_inputs$seed,
  mrmega_methods = "MRMEGA_GEM",
  mrmega_pc_num = pc_num_bin
)

binary_summary
```

